---
title: "Seattle Crimes EDA"
date: '2021-07-01T21:42:38-07:00'
tags:
- Analytics
- Storytelling
comments: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Context

Year 2020 was full of surprises, especially in Seattle. The historical cancellation of the New Year's fireworks seemed to be the premonition of the year to come. Grasped with the pandemic and social unrest, headlines of frequent crimes along with the police defunding movement, it makes one question if crime is actually on the rise in Seattle. This became the inspiration for this project.

### How I did this

This notebook is an in-depth analysis of Seattle Crimes using:
- Seattle Police Department (SPD) Crimes Data downloaded from [Seattle Open Data](https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5) on June 30th 2021. 

- Population estimate for Seattle City from the US Census Bureau for [2000-2010](https://www.census.gov/data/datasets/time-series/demo/popest/intercensal-2000-2010-cities-and-towns.html) and [2010-2019](https://www.census.gov/data/datasets/time-series/demo/popest/2010s-total-cities-and-towns.html) downloaded on June 30th 2021.
  + Population estimates for 2020 were copied from 2019 as no new data is available yet.


```{r, echo=FALSE,message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)
library(reshape2)
library(kableExtra)
library(gridExtra)
library(knitr)
```

```{r, echo=FALSE, message=FALSE}
# load csv file.
crimes = read.csv(file = "SEACrimes_2008_Present.csv")
census00.10 = read.csv(file = "pop2000-2010.csv")
census10.19 = read_excel("H:\\Lily\\Documents\\Portfolio\\website\\content\\project\\1 - SeattleCrimes\\pop2010-2019.xlsx", skip = 3)
names(census10.19)[names(census10.19) == "...1"] = "City.Name"
```

```{r, echo=FALSE, message=FALSE}
# Format date and time for later data extraction.
crimes$DateTime = as.POSIXct(as.character(crimes$Offense.Start.DateTime), format =  "%m/%d/%Y %H:%M")
crimes$Date = as.Date(crimes$DateTime)
crimes = separate(crimes, Date, c("Year", "Month", "Day"), sep = "[-]")

crimes$Time = format(crimes$DateTime, "%T")
crimes = separate(crimes, Time, c("Hour", "Min", "Sec"), sep = "[:]")
```


```{r, echo=FALSE}
# Remove 2021 data, only taking complete year reports and removed crimes occurred before 2008, due to very small numbers (post-dated reporting)
crimes.pre2021 = filter(crimes, Year <=2020 & Year >=2008)

crimes.pre2021.year = crimes.pre2021 %>% 
  group_by(Year) %>% 
  summarise(count=n()) %>% 
  mutate(percent = count/sum(count))
```

#### Crime rate per year:

If we examine the occurrence of crimes each year (2008-2020, 2021 removed due to partial data) as a fraction of total crimes, we get the following:

```{r, echo=FALSE}
# Create new theme for plot
plot.theme1 = theme(
  plot.title = element_text(size = 10, hjust = 0.5),
  axis.title.x = element_text(size = 10),
  axis.text.x = element_text(angle=45, hjust=1, size = 8),
  axis.title.y = element_text(size = 10),
  axis.text.y = element_text(size = 8))
```


```{r, echo=FALSE}
# extract pop estimate for years 2008 and 2009
census00.10.melt = melt(census00.10, id = c(1:7))

pop08.09 = unique((filter(census00.10.melt, (NAME == "Seattle city" & (variable == "POPESTIMATE2008" | variable == "POPESTIMATE2009"))))$value)

# loop to add pop values to new vector for later column addition
pop.by.year = vector("integer", 13L)
for (i in 1:2)
  pop.by.year[i] = pop08.09[i]

# extract pop estimate for years 2010 to 2020.
census10.19.melt = melt(census10.19, id = c(1:3))

pop10.19 = filter(census10.19.melt, City.Name == "Seattle city, Washington")$value

# loop to add pop values to new vector for later column addition
for (i in 3:12)
  pop.by.year[i] = pop10.19[i-2]

# add 2019 pop value for 2020 as new data not available yet.
pop.by.year[13] = pop10.19[10]

# add pop vector to year crime count table to compute crime rate per 100000 pop
crimes.pre2021.year["population"] = pop.by.year

# compute total crime rate per 100000 pop
crimes.pre2021.year = mutate(crimes.pre2021.year,
                             crime.rate = (count/population)*100000)

# plot of fraction of crime per year
plot.fraction = ggplot(data = crimes.pre2021.year, mapping=aes(x=Year, y=percent)) + 
  geom_point() +
  geom_line(group = 1) +
  labs(x = "Year", y = "Fraction of crimes committed", title = "Percentage of crimes committed each year") +
  plot.theme1

# plot of crime per pop per year
plot.rate = ggplot(data = crimes.pre2021.year, mapping=aes(x=Year, y=crime.rate)) + 
  geom_point() +
  geom_line(group = 1) +
  labs(x = "Year", y = "Crime rate per 100,000 pop", title = "Crime rate per 100,000 population per year") +
  plot.theme1

grid.arrange(plot.fraction, plot.rate, nrow=1, ncol=2)

```

A simple plot (*on the left*) of crime counts per year over total crimes of all years (2021 removed due to partial data) shows that crimes drastically decreased in the years of 2011-2012, then steadily increased with the highest counts in 2020. There is a small dip in that can be noted in 2019. Therefore, it seems at first glance that after 2012, crimes increased in Seattle.

However, a more telling plot of crime rate per 100,000 population (*on the right*) can help us examine our assumptions further. In fact, we can see that crime rate is actually very high in 2008 and 2009, and steadily decreased to the lowest level in 2012. Surprisingly, the small dip observed on the previous plot in 2019 revealed a more significant decline in crime rate per population. Last thing to note is that there is a increase of crime in 2020 that can be observed in both plots.

Thus, let's examine the differences between year 2019 and 2020.

```{r, fig.align="center", fig.height = 6, fig.width = 8, echo=FALSE, message=FALSE}

year.pre2021 = crimes.pre2021 %>% 
  group_by(Offense, Year) %>% 
  summarise(count=n())

# select only year 2019 and 2020
year.19.20 = filter(year.pre2021, Year ==2020 | Year ==2019)

# find out top 5 crimes for both years
year.19.20.merge = merge(year.19.20, crimes.pre2021.year, by = "Year", all.x = T)

year.19.20.merge = mutate(year.19.20.merge,
                          crime.rate.ind = (count.x/population)*100000)

year.19.20.order = year.19.20.merge[order(-year.19.20.merge$crime.rate.ind, year.19.20.merge$Offense),]

# Compute top 10 type of offenses and store in a vector
x = unique(year.19.20.order$Offense)
y = x[1:10]

#subset top 10 offenses for 2019 and 2020
year.19.20.subset = year.19.20.merge[year.19.20.merge$Offense %in% y,]

ggplot(year.19.20.subset, aes(fill=Year, y=crime.rate.ind, x=Offense)) +
  geom_bar(position="dodge", stat="identity") +
  labs(x = "Type of Offense", y = "Crime rate per 100,000 population", title = "Crime rate of top 10 Offenses in 2019 and 2020") + 
  theme(plot.title = element_text(hjust = 0.5), axis.text.x=element_text(angle=45,hjust=1))

```

#### 2019 vs 2020

##### Offense Types:

* **Identity Theft** is drastically high in 2020. Could the pandemic be a fertile ground for such activities given the new shift of work from home, unemployment benefits and online shopping? Coincidentally, this finding correlates with multiple reports which presents similar phenomenon in other communities throughout the country, such as [Florida](https://www.aarp.org/money/scams-fraud/info-2021/ftc-fraud-report-identity-theft-pandemic.html) or [California](https://sacramento.cbslocal.com/2020/09/15/edd-putting-people-risk-identity-theft/).

+ It should also be noted that the Seattle identity theft rate of **1273/100,000** is almost double than the Washington state rate of [**712/100,000**](https://www.iii.org/fact-statistic/facts-statistics-identity-theft-and-cybercrime)!

* **Theft from Motor Vehicle** decreased from 2019. One could guess that remote working and unemployment resulted in less vehicles parked in the streets during work hours.

* **Trespass of Real Property** also decreased from 2019, which could be from the same reasons as above. 

* **Shoplifting** also shows a decline in 2020. One can speculate that the closure of stores during the lockdown made it difficult for this type of crime to occur.

* **Burglary/Breaking & Entering** increased in 2020. At first it may be counter-intuitive as one assumes people are at home all the time. It leads to the following question, are the location residences or commercial? Maybe commercial properties that are vacated during the pandemic became a prime target?

##### Where is the crime?

```{r, echo=F, message = FALSE}
MCPP.19.20 = crimes %>% 
  filter(Year ==2020 | Year ==2019) %>% 
  group_by(MCPP, Year) %>% 
  summarise(count=n())

MCPP.19.20.order = MCPP.19.20[order(-MCPP.19.20$count, MCPP.19.20$MCPP),]

# Compute top 10 type of offenses and store in a vector
a = unique(MCPP.19.20.order$MCPP)
b = a[1:10]

#subset top 10 offenses for 2019 and 2020
MCPP.19.20.subset = MCPP.19.20[MCPP.19.20$MCPP %in% b, ]

ggplot(MCPP.19.20.subset, aes(fill=Year, y=count, x=MCPP)) +
  geom_bar(position="dodge", stat="identity") +
  labs(x = "MCPP neighborhoods", y = "Crime count", title = "Top 10 MCPP neighborhoods for crime occurrences in 2019 and 2020") + 
  theme(plot.title = element_text(hjust = 0.5), axis.text.x=element_text(angle=45,hjust=1))
```

There is a drastic decrease in crimes in the Downtown Commercial neighborhood in 2020, and a noticeable increase in Sandpoint and Queen Anne neighborhoods.
Could this be attributed to increase police presence in Downtown? OR the opposite, the decrease of police presence in Sandpoint and Queen Anne?

##### Downtown Crimes by Hour

```{r, echo=FALSE}
with(subset(crimes, MCPP == "DOWNTOWN COMMERCIAL"), qqplot(x = Hour[ Year == '2020'], y = Hour[ Year == '2019'], main='QQ plot for Downtown Crimes'))
abline(0,1)
```

Above QQ plot of hourly crimes between 2019 and 2020 in MCPP Downtown show that the trend is quite different for both years. Crimes tend to happen later in terms of hours in 2019 compared to 2020. The trend tends to become similar for the late afternoon and night hours.


#### 2020

##### Top Crime locations

Remember from the last part where we hypothesized that burglary and breaking crimes increased in commercial addresses in 2020? However, at the same time, total crimes decreased in the Downtown neighborhood which seems contradictory. Let's verify what is going on!

```{r, echo=F, message = FALSE}
crimes.2020 = filter(crimes, Year == 2020)

# filter offense to only burglary and remove address where the value is empty
crimes.2020.burglary = filter(crimes.2020, Offense == "Burglary/Breaking & Entering", Longitude != 0, Latitude != 0)

crimes.2020.burglary.coord = crimes.2020.burglary %>% 
  group_by(Longitude, Latitude) %>% 
  summarise(count=n())

crimes.2020.burglary.coord = crimes.2020.burglary.coord[order(-crimes.2020.burglary.coord$count),]

# take the top 5 addresses
crimes.2020.burglary.top = head(crimes.2020.burglary.coord, 5)
rownames(crimes.2020.burglary.top) = c()

# add manually addresses from Google search coordinates
crimes.2020.burglary.top$address = c("1255 Harrison Street",
                                     "1410 2nd Avenue",
                                     "539 Northeast Northgate Way",
                                     "833 Northeast 67th Street",
                                     "528 Broadway East")

# Display data frame in a table.
crimes.2020.burglary.top %>%
  kable(caption = "Top 5 coordinates-addresses with the most Burglary/Breaking crimes in 2020", col.names = c("Longitude", "Latitude", "Crime Count", "Address")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
* **1255 Harrison Street**: is actually a newly built mixed-use housing and retail condo building.
* **1410 2nd Avenue**: is a Target store in downtown area.
* **539 Northeast Northgate Way**: does not generate an existing address but the closest address, 540, is a Starbucks. Surrounding buildings are all commercial on the ground floor.
* **833 Northeast 67th Street**: empty lot in construction.
* **528 Broadway East** : 527 is a restaurant, surrounding buildings are all commercial on the ground floor.

Although *one should keep in mind that the addresses generated are an approximation based on the Latitude and Longitude coordinates*, above table gives a pretty clear picture of where burglary tend to occur. Our hypothesis seem to be holding, commercial properties are most likely the target for burglary and breaking crimes.


##### Prime Crime hours

Do different crimes have different occurrence patterns throughout the day?

```{r, fig.align="center", fig.height = 8, fig.width = 8, echo=FALSE, message=FALSE}

# select only year 2020
year.20 = filter(year.pre2021, Year ==2020)

year.20.order = year.20[order(-year.20$count), ]

# Compute top 10 type of offenses and store in a vector
c = unique(year.20.order$Offense)
d = c[1:10]

#subset top 10 offenses for 2019 and 2020
year.20.top = year.20[year.20$Offense %in% d,]

year.20.merge = merge(crimes.2020, year.20.top, by = "Offense")

hour.2020 = crimes.2020 %>% 
  group_by(Hour) %>% 
  summarise(count=n()) %>% 
  mutate(percent = count/sum(count))

hour.2020.offense = year.20.merge %>% 
  group_by(Offense, Hour) %>% 
  summarise(count=n()) %>% 
  mutate(percent = count/sum(count))

ggplot(data = hour.2020.offense, mapping=aes(x=Hour, y=percent)) + 
  geom_point(size=0.5) + geom_line(color = "blue", group =1) + facet_wrap("Offense", ncol = 2) + 
  geom_line(data = hour.2020, mapping=aes(x=Hour, y=percent), color = "grey", group=1) +
  labs(x = "Hour", y = "Fraction of crimes committed", title = "Percentage of crimes committed each hour by offense type") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
   scale_x_discrete(breaks = c("00", "04", "08", "12", "16", "20", "24"), labels = c("0", "4", "8", "12", "16", "20", "24"))
```

Compared to the overall crime occurrence trend line (in gray), a few observations stand out:

- **Midnight** is when the most crimes occur.

- **Noon** is when many crimes see a small peak.

- *Identity theft* experiences a huge peak around midnight and is high in the morning.
  + If we think about the mechanism of this crime, the exact moment the crime occurs may be unknown and when people reported it, an arbitrary time may have be chosen, which may explain the peak around midnight.
  + Morning hours may likely be when people realize and report that they experienced identity theft.
  
- *Shoplifting* occurs mostly during daytime, which correlates with commercial opening hours.

- *Theft-type* of crimes all experience a small increase in the evening, all with peak around midnight.
  + Again, one may question if the crime occurred around midnight or due to the delayed discovery of the crime, an arbitrary time was chosen.


### Conclusion

This was a brief overview of some interesting aspects of the Seattle Crime dataset. From above analysis, we can conclude that:

- Crime rate increased in 2020 from 2019.
- The pandemic may have played a role in the rise of Identity Theft while deterring Shoplifting and Theft from Motor Vehicle.
- Seattle experienced a decrease in crimes in the Downtown area, although burglary still commonly occur in commercial settings.
- Midnight seem to be the time of day with the highest crime occurrences, which may or may not be contributed to systemic data entering. Noon is also a peak time for crimes.


** This project's repository can be found on my [github](https://github.com/lilyportfolio/website/tree/main/content/project/1%20-%20SeattleCrimes) **