---
title: "Project 1 - Seattle Crime"
date: 2021-07-01T21:42:38-07:00
tags: [Analytics]
comments: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

This notebook is an in-depth analysis of Seattle Police Department (SPD) Crimes Data downloaded from [Seattle Open Data](https://data.seattle.gov/Public-Safety/SPD-Crime-Data-2008-Present/tazs-3rd5) on June 30th 2021. 


```{r, echo=FALSE,message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r, echo=FALSE}
# load csv file.
crimes = read.csv(file = "SEACrimes_2008_Present.csv")
```

```{r, echo=FALSE}
# Format date and time for later data extraction.
crimes$Report.DateTime = as.POSIXct(as.character(crimes$Report.DateTime), format =  "%m/%d/%Y %H:%M")
crimes$Report.Date = as.Date(crimes$Report.DateTime)
crimes = separate(crimes, Report.Date, c("Report.Year", "Report.Month", "Report.Day"), sep = "[-]")

crimes$Report.Time = format(crimes$Report.DateTime, "%T")
crimes = separate(crimes, Report.Time, c("Report.Hour", "Report.Min", "Report.Sec"), sep = "[:]")
```

If we examine the occurrence of crimes each year (2008-2020, 2021 removed due to partial data) as a fraction of total crimes, we get the following plot:

```{r, echo=FALSE}
# Remove 2021 data, only taking complete year reports
crimes.pre2021 = filter(crimes, Report.Year <=2020)

crimes.pre2021.year = crimes.pre2021 %>% 
  group_by(Report.Year) %>% 
  summarise(count=n()) %>% 
  mutate(percent = count/sum(count))

ggplot(data = crimes.pre2021.year, mapping=aes(x=Report.Year, y=percent)) + 
  geom_point() +
  geom_line(group = 1) +
  labs(x = "Year", y = "Fraction of crimes committed", title = "Percentage of crimes committed each year") +
  theme(plot.title = element_text(hjust = 0.5))
```

Above plot shows that crime rates in Seattle decreased in the year *2011-2012*, then steadily increases starting in 2013, with the highest rate in *2020*. That could be interesting. Let's compare these two years in terms of type of Offenses to see if we can detect any differences.

```{r, fig.align="center", fig.height = 8, fig.width = 8, echo=FALSE, message=FALSE}

year.pre2021 = crimes.pre2021 %>% 
  group_by(Offense, Report.Year) %>% 
  summarise(count=n())
year.pre2021= mutate(year.pre2021,
                     percent = count/sum(year.pre2021$count))

# Create a reference data frame with all crimes in each year for later comparison.
year.no.facet = subset(year.pre2021, select = c("percent", "Offense"))

year.12.20 = filter(year.pre2021, Report.Year ==2020 | Report.Year ==2012)

# Create plot showing the percentage of crimes committed in 2012 and 2020 respective to all crimes in the same year
ggplot(data = year.12.20, mapping=aes(x = percent, y = Offense)) + 
  geom_point(data = year.no.facet, color="grey", size = 1.5) +
  geom_point(size = 1.0) + facet_wrap(~Report.Year) +
  labs(x = "Fraction of crimes committed", y = "Crimes", title = "Fraction of crimes committed in 2012 and 2020 by offense type") + theme(plot.title = element_text(hjust = 0))

```

Interestingly (or not), other than **Identity Theft** which occurred at at the highest rate in 2020 compared to all previous years, the other most occurring crimes (in terms of fraction compared to total crimes in the year) seem to happen at a fairly similar rate. Those crimes include:

- Theft from Motor Vehicle
- Simple Assault
- Shoplifting
- Motor Vehicle Theft
- Destruction/Damage/Vandalism of Property
- Burglary/Breaking & Entering
- All Other Larceny
- Aggravated Assault

Let's then examine those crimes in the year 2020, which remember, was the year with the highest crime occurrences.
First, do the crimes differ in terms of hour of the day?

```{r, fig.align="center", fig.height = 8, fig.width = 8, echo=FALSE, message=FALSE}

crimes.2020 = filter(crimes, Report.Year == 2020)

crimes.2020.offense = filter(crimes.2020, Offense == "Theft from Motor Vehicle" | Offense == "Simple Assault" | Offense == "Motor Vehicle Theft" | Offense == "Identity Theft" | Offense == "Destruction/Damage/Vandalism of Property" | Offense == "Burglary/Breaking & Entering" | Offense == "All Other Larceny")

hour.2020 = crimes.2020.offense %>% 
  group_by(Report.Hour) %>% 
  summarise(count=n()) %>% 
  mutate(percent = count/sum(count))

hour.2020.offense = crimes.2020.offense %>% 
  group_by(Offense, Report.Hour) %>% 
  summarise(count=n()) %>% 
  mutate(percent = count/sum(count))

ggplot(data = hour.2020.offense, mapping=aes(x=Report.Hour, y=percent)) + 
  geom_point(size=0.5) + geom_line(color = "blue", group =1) + facet_wrap("Offense", ncol = 2) + 
  geom_line(data = hour.2020, mapping=aes(x=Report.Hour, y=percent), color = "grey", group=1) +
  labs(x = "Hour", y = "Fraction of crimes committed", title = "Percentage of crimes committed each hour by offense type") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
   scale_x_discrete(breaks = c("00", "04", "08", "12", "16", "20", "24"), labels = c("0", "4", "8", "12", "16", "20", "24"))
```

Compared to the overall crime occurrence trend line (in gray), a few observations stand out:

- *Identity theft* is very high during working hours and remains after 16:00.
  + It could be that only during wake hours people realized that they experienced identity theft.
  + It could be that this kind of crime do not require immediate attention and could be deferred for reporting during working hours.
  
- *Simple Assault*